### 2024-09-17 base-rates.qmd
20240914 - base rates
20240916 - Nate Silver examples

@@

The [`petersenlab`](https://github.com/DevPsyLab/petersenlab) package [@R-petersenlab] contains the `pAgivenB()` function that estimates the probability of one event, $A$, given another event, $B$.

@@

@@

We can compute the term in the denominator ($P(\text{Identified as Blue})$) using the law of total probability (described in @sec-bayesTheorem).
$$
\begin{aligned}
  P(R_i) &= P(R | C) \cdot P(C_i) + P(R | \text{not } C) \cdot P(\text{not } C_i)\\
  P(R_i) &= P(\text{Identified as Blue}|\text{Blue}) \cdot P(\text{Blue}) + P(\text{Identified as Blue}|\text{Green}) \cdot P(\text{Green})\\
  0.29 &= (.80 \times .15) + (.20 \times .85) \\
\end{aligned}
$$ {#eq-cabExampleLawOfTotalProbability}

@@

@@

Base rates are also relevant to fantasy football.
Unlike yardage (e.g., passing yards, rushing yards, receiving yards), touchdowns occur *relatively* less frequently.
Whereas a solid Wide Receiver may log 100+ receptions and 1,200+ yards in a season, they may have "only" 8–14 receiving touchdowns in a given season.
As noted in @sec-differentErrorsDifferentCosts, lower base-rate events—including touchdowns—are harder to predict accurately.
As noted by [Harris (2012)](https://www.espn.com/fantasy/football/ffl/story?page=nfldk2k12_vbdwork): "NFL statistical projections are basically impossible to get right. (Take it from someone who helps create them for a living.) Yes, we can do a passable job with yardage totals for players who don't suffer unexpected injuries or depth-chart pratfalls. But so much of fantasy football hinges on touchdowns, and touchdowns are impossibly difficult to predict from season to season (let alone week to week)." (archived at <https://perma.cc/4QNH-J2LD>).
Thus, it is important not to lend too much credence to predictions of touchdowns.
Focus on other things that may be more predictable (and that may be indirectly prognostic of touchdowns) such as yards, carries/targets, receptions, depth of targets, red zone carries/targets, short distance carries/targets, etc.
When dealing with numeric predictions (rather than categorical outcomes), the equivalent of the base rate is the average value.
For instance, the "base rate" of fantasy points for a given position is the average number of fantasy points for that position.

@@

>>>>

#### Standard Formulation {#sec-bayesTheoremStandard}

@@

@@

#### Alternative Formulation {#sec-bayesTheoremAlternative}

Using the law of total probability, we can substitute the calculation of the [marginal probability](#sec-baseRate) ($P(R_i)$) into Bayes' theorem to get an alternative formulation of Bayes' theorem, as in @eq-baseRateUpdated3:

$$
\begin{aligned}
  P(C | R) &= \frac{P(R | C) \cdot P(C_i)}{P(R_i)} \\
  &= \frac{P(R | C) \cdot P(C_i)}{P(R | C) \cdot P(C_i) + P(R | \text{not } C) \cdot P(\text{not } C_i)} \\
  &= \frac{P(R | C) \cdot P(C_i)}{P(R | C) \cdot P(C_i) + P(R | \text{not } C) \cdot [1 - P(C_i)]}
\end{aligned}
$$ {#eq-baseRateUpdated3}

Instead of using [marginal probability](#sec-baseRate) ([base rate](#sec-baseRate)) of $R$, as in the original formulation of Bayes' theorem, it uses the [conditional probability](#sec-conditionalProbability), $P(R|\text{not } C)$.
Thus, it uses three terms: two [conditional probabilities](#sec-conditionalProbability)—$P(R|C)$ and $P(R|\text{not } C)$—and one [marginal probability](#sec-baseRate), $P(C_i)$.

Let us see how the alternative formulation of Bayes' theorem applies to the HIV example above.
We can calculate the probability of HIV given a positive test using three terms: the [conditional probability](#sec-conditionalProbability) that a person with HIV gets a positive test (i.e., [sensitivity](#sec-sensitivity)), the [conditional probability](#sec-conditionalProbability) that a person without HIV gets a positive test (i.e., [false positive rate](#sec-falsePositiveRate)), and the [base rate](#sec-baseRate) of HIV.
Using the $P(R|\text{not } C)$ calculated in @eq-conditionalProbabilityRevised, the [conditional probability](#sec-conditionalProbability) of HIV given a single positive test is in @eq-bayes4:

$$
\small
\begin{aligned}
  P(C | R) &= \frac{P(R | C) \cdot P(C_i)}{P(R | C) \cdot P(C_i) + P(R | \text{not } C) \cdot [1 - P(C_i)]} \\
  &= \frac{\text{sensitivity of test} \times \text{base rate of HIV}}{\text{sensitivity of test} \times \text{base rate of HIV} + \text{false positive rate of test} \times (1 - \text{base rate of HIV})} \\
  &= \frac{95\% \times .3\%}{95\% \times .3\% + .7171515\% \times (1 - .3\%)} = \frac{.95 \times .003}{.95 \times .003 + .007171515 \times (1 - .003)}\\
  &= 28.5\%
\end{aligned}
$$ {#eq-bayes4}

```{r}
pAgivenB(
  pBgivenA = .95,
  pA = .003,
  pBgivenNotA = .007171515)

pAgivenB(
  pBgivenA = .95,
  pA = .003,
  pBgivenNotA = pBgivenNotA(
    pBgivenA = .95,
    pA = .003,
    pB = .01))
```

To calculate the [conditional probability](#sec-conditionalProbability) of HIV given a second positive test, we update our priors because the person has now tested positive for HIV.
We update the [prior probability](#sec-baseRate) of HIV ($P(C_i)$) based on the posterior probability of HIV after a positive test ($P(C | R)$) that we calculated above.
We can calculate the [conditional probability](#sec-conditionalProbability) of HIV given a second positive test using three terms: the [conditional probability](#sec-conditionalProbability) that a person with HIV gets a positive test (i.e., [sensitivity](#sec-sensitivity); which stays the same), the [conditional probability](#sec-conditionalProbability) that a person without HIV gets a positive test (i.e., [false positive rate](#sec-falsePositiveRate); which stays the same), and the updated [marginal probability](#sec-baseRate) of HIV.
The [conditional probability](#sec-conditionalProbability) of HIV given a second positive test is in @eq-baseRateUpdated4:

$$
\scriptsize
\begin{aligned}
  P(C | R) &= \frac{P(R | C) \cdot P(C_i)}{P(R | C) \cdot P(C_i) + P(R | \text{not } C) \cdot [1 - P(C_i)]} \\
  &= \frac{\text{sensitivity of test} \times \text{updated base rate of HIV}}{\text{sensitivity of test} \times \text{updated base rate of HIV} + \text{false positive rate of test} \times (1 - \text{updated base rate of HIV})} \\
  &= \frac{95\% \times 28.5\%}{95\% \times 28.5\% + .7171515\% \times (1 - 28.5\%)} = \frac{.95 \times .285}{.95 \times .285 + .007171515 \times (1 - .285)}\\
  &= 98.14\%
\end{aligned}
$$ {#eq-baseRateUpdated4}

```{r}
pAgivenB(
  pBgivenA = .95,
  pA = .285,
  pBgivenNotA = .007171515)

pAgivenB(
  pBgivenA = .95,
  pA = .285,
  pBgivenNotA = pBgivenNotA(
    pBgivenA = .95,
    pA = .003,
    pB = .01))
```

#### Interim Summary

@@

We can compute the term in the denominator ($P(\text{Identified as Blue})$) using the law of total probability (described in @sec-bayesTheorem).

$$
\begin{aligned}
  P(R_i) &= P(R | C) \cdot P(C_i) + P(R | \text{not } C) \cdot P(\text{not } C_i)\\
  P(R_i) &= P(\text{Identified as Blue}|\text{Blue}) \cdot P(\text{Blue}) + P(\text{Identified as Blue}|\text{Green}) \cdot P(\text{Green})\\
  0.29 &= (.80 \times .15) + (.20 \times .85) \\
\end{aligned}
$$ {#eq-cabExampleLawOfTotalProbability}

@@

## Nate Silver Examples {#sec-nateSilverExamples}

@Silver2012 provides several examples that leverage the [alternative formulation of Bayes' theorem](#sec-bayesTheoremAlternative) provided in @eq-baseRateUpdated3 and summarized below:

$$
\begin{aligned}
  P(C | R) &= \frac{P(R | C) \cdot P(C_i)}{P(R | C) \cdot P(C_i) + P(R | \text{not } C) \cdot [1 - P(C_i)]}
\end{aligned}
$$ {#eq-baseRateUpdated3}

In each case, the formula uses three elements to calculate the probability that the hypothesis is true:

1. the [conditional probability](#sec-conditionalProbability) the likelihood of observing the evidence, $R$, given that the hypothesis, $C$, is true (i.e., $P(R|C)$; [sensitivity](#sec-sensitivity))
1. the [conditional probability](#sec-conditionalProbability) the likelihood of observing the evidence, $R$, given that the hypothesis, $C$, is false (i.e., $P(R | \text{not } C)$; [false positive rate](#sec-falsePositiveRate))
1. the [marginal probability](#sec-baseRate) (base rate) of the event occurring (i.e., the prior probability of the hypothesis, $C$ being true; $P(C_i)$)

Example 1: You just came home and a strange pair of underwear in your drawer.
What is the probability that your partner is cheating on you?

- the [prior probability](#sec-baseRate) that your spouse is cheating on you: 4%
- the [conditional probability](#sec-conditionalProbability) of underwear appearing given that your spouse is cheating on you: 50%
- the [conditional probability](#sec-conditionalProbability) of underwear appearing given that your spouse is *not* cheating on you: 5%

```{r}
pAgivenB(
  pBgivenA = .50,
  pA = .04,
  pBgivenNotA = .05)
```

Example 2: What is the probability that a woman in her 40s has breast cancer if she tested positive on a mammogram?

- the [prior probability](#sec-baseRate) that she has cancer: 1.4%
- the [conditional probability](#sec-conditionalProbability) that she has a positive test given that she has cancer: 75%
- the [conditional probability](#sec-conditionalProbability) that she has a positive test given that she does *not* have cancer: 10%

```{r}
pAgivenB(
  pBgivenA = .75,
  pA = .014,
  pBgivenNotA = .10)
```

Example 3A: Consider the information we had on 9/11 when the first plane hit the World Trade Center.
What is the probability that a terror attack occurred given that the first plane hit the World Trade Center?

- the [prior probability](#sec-baseRate) that terrorists crash a plane into a Manhattan skyscraper: 0.005%
- the [conditional probability](#sec-conditionalProbability) that a plane crashes into a Manhattan skyscraper if terrorists are attacking Manhattan skyscrapers: 100%
- the [conditional probability](#sec-conditionalProbability) that a plane crashes into a Manhattan skyscraper if terrorists are *not* attacking Manhattan skyscrapers (i.e., it is an accident): 0.008%

```{r}
pAgivenB(
  pBgivenA = 1,
  pA = .00005,
  pBgivenNotA = .00008)
```

Example 3A: Now, consider that a second plan just hit the World Trade Center.
What is the probability that a terror attack occurred given that a second plane hit the World Trade Center?

- the revised [prior probability](#sec-baseRate) that terrorists crash a plane into a Manhattan skyscraper (from Example 3A): 38.46272%
- the [conditional probability](#sec-conditionalProbability) that a plane crashes into a Manhattan skyscraper if terrorists are attacking Manhattan skyscrapers: 100%
- the [conditional probability](#sec-conditionalProbability) that a plane crashes into a Manhattan skyscraper if terrorists are *not* attacking Manhattan skyscrapers (i.e., it is an accident): 0.008%

```{r}
pAgivenB(
  pBgivenA = 1,
  pA = .3846272,
  pBgivenNotA = .00008)
```

@@

Base rates are also relevant to fantasy football.
Unlike yardage (e.g., passing yards, rushing yards, receiving yards), touchdowns occur *relatively* less frequently.
Whereas a solid Wide Receiver may log 100+ receptions and 1,200+ yards in a season, they may have "only" 8–14 receiving touchdowns in a given season.
As noted in @sec-differentErrorsDifferentCosts, lower base-rate events—including touchdowns—are harder to predict accurately.
As noted by [Harris (2012)](https://www.espn.com/fantasy/football/ffl/story?page=nfldk2k12_vbdwork): "NFL statistical projections are basically impossible to get right. (Take it from someone who helps create them for a living.) Yes, we can do a passable job with yardage totals for players who don't suffer unexpected injuries or depth-chart pratfalls. But so much of fantasy football hinges on touchdowns, and touchdowns are impossibly difficult to predict from season to season (let alone week to week)." (archived at <https://perma.cc/4QNH-J2LD>).
Thus, it is important not to lend too much credence to predictions of touchdowns.
Focus on other things that may be more predictable (and that may be indirectly prognostic of touchdowns) such as yards, carries/targets, receptions, depth of targets, red zone carries/targets, short distance carries/targets, etc.

@@

When dealing with numeric predictions (rather than categorical outcomes), the equivalent of the base rate is the average value.
For instance, the "base rate" of fantasy points for a given position is the average number of fantasy points for that position.
We could subdivide even further to identify, for instance, the "base rate" of fantasy points for the Wide Receiver at the top of the depth chart on a team.
